---
layout: post
title: "Getting to Know Your Data"
date:   2015-10-22 20:49:52
authors: "Courtney Soderberg"
dateCreated:  2014-11-26 20:49:52
lastModified: 2015-07-23 14:28:52
category: time-series-workshop
tags: [module-1]
mainTag: GIS-Spatial-Data
description: "This post explains the fundamental principles, functions and metadata that you need to work with raster data in R."
image:
  feature: lidar_GrandMesa.png
  credit: LiDAR data collected over Grand Mesa, Colorado - National Ecological Observatory Network (NEON)
  creditlink: http://www.neoninc.org
permalink: /R/Raster-Data-In-R/
code1: /R/2015-07-22-Introduction-to-Raster-Data-In-R.R
comments: true

---

<section id="table-of-contents" class="toc">
  <header>
    <h3>Contents</h3>
  </header>
<div id="drawer" markdown="1">
*  Auto generated table of contents
{:toc}
</div>
</section><!-- /#table-of-contents -->

##About
This activity will walk you through the fundamentals of data manipulation and basic plotting.

**R Skill Level:** Intermediate - you've got the basics of `R` down.

<div id="objectives">

<h3>Goals / Objectives</h3>
After completing this activity, you will know:
<ol>
<li>How to calculate basic descriptive statistics for groups of data.</li>
<li>How to create basic time series plots in `R`.</li>
</ol>

<h3>Things You'll Need To Complete This Lesson</h3>

<h3>R Libraries to Install:</h3>
<ul>
<li><strong>raster:</strong> <code> install.packages("dplyr")</code></li>
<li><strong>rgdal:</strong> <code> install.packages("lubridate")</code></li>
<li><strong>rgdal:</strong> <code> install.packages("ggplot2")</code></li>

</ul>
<h4>Tools To Install</h4>

Please be sure you have the most current version of `R` and preferably
R studio to write your code.

#Plotting Time Series Data
One of the first things that it can often be useful to do once we've loaded our data and cleaned it up is to visualize our data. We can start to get a sense of general trends, and well as see possible outliers or non-sensical values. 

To do this, we're going to use `ggplot2` to plot the air temperature across our 2 year span time span for each 15 minute data point.

```{r basic plotting}
#plot Some Air Temperature Data
myPlot <- ggplot(yr.09.11,aes(datetime, airt)) +
           geom_point() +
           ggtitle("15 min Avg Air Temperature\nHarvard Forest") +
           theme(plot.title = element_text(lineheight=.8, face="bold",size = 20)) +
           theme(text = element_text(size=20)) +
           xlab("Time") + ylab("Mean Air Temperature")

#format x axis with dates
myPlot + scale_x_datetime(labels = date_format("%m/%d/%y"))
```

#Manipulating Data
This plot is interesting, and we can see that there is a clear seasonal trend, unsurprisingly. But the air temperature data at 15 minute intervals might be more granular than we want. Instead, we may want to look at how daily average temperature changes over time. To do this we first need to learn a bit about how to manipulate data in R. We'll use the `dplyr` package and a "split-apply-combine" technique.

We use the `group-by()` function to determine how we split up our data. For example, if we wanted to count up the number of observations per Julian dat, we would do:

```{r dplyr group}
yr.09.11 %>%
  group_by(jd) %>%
  tally()
```

We're more interested though in calculting single values by julian day. We can use the `summarize()` function for this. To get the mean air temperature for each julian day, we would write:
```{r dplyr summarize}
yr.09.11 %>%
  group_by(jd) %>%
  summarize(mean_airt = mean(airt, na.rm = TRUE))
```

Julian days repeat though, so if we want to summarize our air temperature data for each day, we actually need to group our data by two different values at once, year and julian day. To do this, we're first going to need to create a new year variable. For this we'll use the `lubridate` package.

```{r dplyr lubridate}
yr.09.11$year <- year(as.Date(yr.09.11$datetime, "%y-%b-%d", tz = "America/New_York"))
```

This code will work, but we previously already set up datetime as a time variable, so we can be a bit more efficient an just put:

```{r dplyr lubridate-2}
yr.09.11$year <- year(yr.09.11$datetime)
```

Challenge: Can you do this for month as well?

Now we have our two variables. So, to get the mean air temperature for each day for each year, we would write:

```{r dplyr full}
yr.09.11 %>%
  group_by(year, jd) %>%
  summarize(mean_airt = mean(airt, na.rm = TRUE))
```

Could we create the year variable within our `dplyr` function call? Yep. We can use the `mutate()` function, and include our `lubridate` function within the `mutate()` call.

```{r dplyr mutate}
yr.09.11 %>%
  mutate(year = year(datetime)) %>%
  group_by(year, jd) %>%
  summarize(mean_airt = mean(airt, na.rm = TRUE))
```

We also want to save this information so that we can plot the daily air temperature as well. So, we save the output of our dplyr function as a new dataframe.
```{r dplyr saving}
temp.daily <- yr.09.11 %>%
  mutate(year = year(datetime)) %>%
  group_by(year, jd) %>%
  summarize(mean_airt = mean(airt, na.rm = TRUE), datetime = first(datetime))
```

Since we're going to plot this, we also want to retain the datetime information. But, we only want one datetime entry per air temperature measurement. So, we ask R to give us the first datetime entry for each day for each year.

Now that we have our new data file, we want to plot the daily averages.

```{r daily plot}
dailyPlot <- ggplot(temp.daily,aes(datetime, mean_airt)) +
           geom_point() +
           ggtitle("Daily Avg Air Temperature\nHarvard Forest") +
           theme(plot.title = element_text(lineheight=.8, face="bold",size = 20)) +
          theme(text = element_text(size=20)) +
           xlab("Time") + ylab("Mean Air Temperature")

#format x axis with dates
dailyPlot + scale_x_date(labels = date_format("%m/%d/%y"))
```