---
layout: post
title: "Lesson 00: Working With Date Formats in R - Intro to Tabular Time Series"
date:   2015-10-25
authors: [Megan A. Jones, Marisa Guarinello, Courtney Soderberg]
contributors: [Leah A. Wasser]
dateCreated:  2015-10-22
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
packagesLibraries: [raster, rgdal]
category: 
tags: [module-1]
description: "This lesson will demonstrate how to import a time series dataset 
stored in .csv format into R. It will explore column formats and will walk
through how to convert a date field, stored as a character string, into a date 
field that R can recogniaze and plot efficiently."
image:
  feature: NEONCarpentryHeader_2.png
  credit: A collaboration between the National Ecological Observatory Network (NEON) and Data Carpentry
  creditlink: http://www.neoninc.org
permalink: /R/Introduction-to-Tabular-Time-Series-In-R
comments: false
---

{% include _toc.html %}

##About
This learning module will walk you through the fundamental principles of working 
with `.csv` temporal data in `R` including how to open, clean, explore, and plot 
time series data. 

<div id="objectives" markdown="1">

**R Skill Level:** Intermediate - you've got the basics of `R` down.

###Goals / Objectives
After completing this learning module, you will:

  * Be able to open a `.csv` file in 'R'using `read.csv`.
  * Understand how to work with `date.frame` fields in `R`.
  * Understand how to examine data structures and types.
  * Be able to convert a date field, stored as a character into an `R` date 
  format.
  * Know how to plot a time-series dataset using `qplot` 

###Things You'll Need To Complete This Lesson
Please be sure you have the most current version of 'R' and, preferably,
RStudio to write your code.

####R Libraries to Install

* **ggplot:** `install.packages("ggplot2")`

 <a href="http://neondataskills.org/R/Packages-In-R/" target="_blank">
More on Packages in R - Adapted from Software Carpentry.</a>

####Data to Download

<a href="http://files.figshare.com/2437700/AtmosData.zip" class="btn btn-success">
Download Atmospheric Data</a>

The data used in this tutorial were collected at Harvard Forest which is
a the National Ecological Observatory Network field site 
<a href="http://www.neoninc.org/science-design/field-sites/harvard-forest" target="_blank">
More about the NEON Harvard Forest field site</a>. These data are proxy data for 
what will be available for 30 years [on the NEON data portal](http://data.neoninc.org/ "NEON data")
for both Harvard forest and other field sites located across the United States.

</div>

#The Data Approach

In this lesson, we will explore micro-meterology data collected from a 
<a href="http://www.neoninc.org/science-design/collection-methods/flux-tower-measurements" target="_blank">flux
tower</a> at the NEON Harvard Forest field site. We are interested in exploring 
patterns associated with changes in temperature, precipitation, PAR and day 
length -- metrics that impact changes in plant phenology 

<link to phenology as a science topic - pbb??>

#About .csv file format

We have received data from a flux tower at Harvard Forest in `.csv` format.
The file extension `.csv` stands for comma separated values. It is a text file 
based format, where each value in the dataset is separate by a comma. Plain text 
formats are ideal for working both across platforms (MAC, PC, LINUX, etc) and 
also can be read by many different tools. Additionally file format is further 
less likely to become obsolete over time!

* <a href="https://en.wikipedia.org/wiki/Comma-separated_values" target=_blank"> For more on .csv format see this Wikipedia article**</a>

Our `.csv` file contains a suite of metrics that we are interested in. To begin 
exploring  the data, we need to first import it into R. We can use the function 
`read.csv` to import the data. We can then explore the data structure.

> Bonus:#Good coding practice -- install and load all libraries at top of script.
> If you decide you need another package later on in the script return to this
> area and add it there.  That way, with a glance, you can see all packages
> used in that script. 

```{r load-libraries}
# Install packages required for entire script.
#install.packages(ggplot2)

# What libraries are required for this lesson?
# library(nameOfLibrary)  #purpose of library
library(ggplot2)   #efficient, pretty plotting - required to qplot function

#set working directory to ensure R can find the file we wish to import
#setwd("workingDirHere")

```

Once our working directory is set, we can import import the data using 
`read.csv`. 

When reading in files we will use `stringsAsFactors = FALSE`. This additon causes
`R` to read non-numeric data (strings) in their original format and not convert 
them to factors.  Factors are non-numerical data that can be numerically 
interpreted and may have a level associated with them. Examples of factors:

* names of the months (an ordinal variable). They are non-numerical but we know 
that *April* comes after *March* and each could be represented by 4 and 3 
respectively.
* 1 and 2s to represent male and female sex (a nominal variable). Numerical interpretation of non-numerical data but no order to the levels.  

But much of our data are not suitable for factors. An example of strings as
non-factors is species observed: *mouse*, *grasshopper*, and *sparrow*; these
are non-numberical data with no relevant level or numerical interpretation.  

It is easy to turn specific colums into factors later on (`as.factor()`), so it 
is almost always best to read in a file with `stringsAsFactors = FALSE`.

```{r import-csv}

#Load csv file of daily meterological data from Harvard Forest
harMetdaily <- read.csv(file="AtmosData/HARV/hf001-06-daily-m.csv",
                     stringsAsFactors = FALSE)

#what type of R object is our imported data?
class(harMetdaily)

```

#Data.Frames in R
When we use the `read.csv()` function, `R` imports the data in `data.frame` 
format. This format is idea for working with tabular data. It is similar to a 
spreadsheet.

## Data Structure & Metadata
Once our data are imported, we can explore its structure. There are several ways to examine the structure of a data frame: 

* `head()` method: this will show us the first 6 rows of the data. 
* `str()` method: view the structure of the data as `R` interprets it.

```{r view-data-structure}

#view first 6 rows of the dataframe 
head(harMetdaily)

#View the structure (str) of the data 
str(harMetdaily)

```

##Data Structure - Formats in R

The structure of our data tells us that there the data are stored in several 
formats in our `.csv`:

* **chr** - is the  character class which holds strings that are data made
up of letters and words.  They are not of a numerical interpretation and most 
often represent data that are discrete classes.  
* **int** - is the integer class which contains 
* **num** - a few sentences.

Storing variables in different formats is a strategic decision by `R`. ... more
on why this is efficient here.

#Plot data using qplot

Let's have a look at one of the metrics, stored within our data - `airt` which is
air temperature. Given this is a time series dataset, we might want to plot
air temperature as it changes over time. We have a date time field so let's use
that as our x axis variable, `airt` will be our Y axis variable.

We will use the `qplot` function in the `ggplot2` library which stands for quick plot.
`qplot` requires the x and y axis variables and then the data that the variables 
are stored in. 

```{r plot-data}
#quickly plot the data
qplot(x=date,y=airt, data=harMetdaily)

```

We have successfully plotted some data. However, what is going on - on the x axis.
R is trying to plot EVERY date value on the x Axis - so it's not readable? Why?
Let's have a look at the format of the fields that we are trying to plot.

#NOTE - We might want to subset this to maybe a single year.

```{r view-field-structure }
# View data structure for each column that we wish to plot
class(harMetdaily$date)

class(harMetdaily$airt)

```

Looking at our data, we can see that the datetime field actually imported into
`R` as a character format. This means that `R` does not know how to plot these 
data as a continuous variable on the X axis. The `airt` field is numeric so that 
field should plot, just fine.

#Working With Time Formats
We need to convert our date field, which is currently stored as a character value
to an actual date format that can be displayed quantitively. Lucky for us, `R` has 
`date` format. We can reassign the `harMetdaily$date` field using `as.date`.

```{r convert-date-time }

#convert field to date format
harMetdaily$date <- as.Date(harMetdaily$date)

#view results
head(harMetdaily$date)

#view structure of field
class(harMetdaily$date)

```

Now that we have adjusted the date, let's plot again. Notice that it plots
much more quickly now that R recognizes the date as a date field. R can agregate
ticks on the x axis by year instead of trying to plot every day!

```{r plot-data}
#quickly plo the data
qplot(x=date,y=airt, 
      data=harMetdaily,
      main="Daily Air Temperature w Date Assigned\nNEON Harvard Forest")
```


#Challenge
> 1. Plot something else?? what else should we plot.
> 2. it could be cool to show them how to aggregate - adjust tick marks by 5 years, 10 years, etc
> 3. it could be cool to show them how to add x lab and y lab
> 4. plot on variable vs another??




############## REMOVE STUFF BELOW ??
#Recommended Pre-Module Reading

## Time series data
Time series data can come in many different forms.  At the fundemental level a 
time series is a series of data points collected at successive time intervals. 
Lots of data collected can be time series data: heights of ocean tides, the 
number of toes that different equine species had over the last 60 mya, or daily 
temperature measures.  This module will specifically look at atmospheric data 
(temperature, precipitation, and photosynthetically active radiation (PAR)) 
recorded over various time intervals (every 1 sec, 15 min, daily) and will also use
the normalized difference vegetative index (NDVI) calculated from remote sensing
data.
 
## PAR
PAR is a measure of how much light within the solar radiation spectral range 
(wave band) from 400 to 700 nanometers is present at the time when the data is 
collected.  This wavelength is important for phenology studies as that spectral
range is what photosynthetic organisms are able to use in the process of 
photosynthesis.

## NDVI
NDVI is an index calculated from spectral reflectance measurements acquired in 
the visible (red) and near-infrared regions by sensors on earth-observing satellites.
This index indicates how green plants in the target area are and is often
used in phenology studies as it can be used to determine growing and dormant seasons.

# Additional information if working through module as an individual, not as part of a workshop. 
This .csv time-series learning module was build as part of the two day Spatio-
Temporal Data Workshop [LINK]XXX. Therefore some background information and 
basic coding skills have been covered extensively in previous modules and are
presented here in a more limited fashion. You should be able to complete this 
module as long as you are comfortable with basic 'R' skills and have a general
understanding of spatial data.   

This workshop is built around Jessica, an ecology graduate student interested in
exploring some of her field sites to understand how phenology of vegetation 
(greening in the spring/summer and senescence in the winter, varies across 
multiple sites and through time). Her potential study sites have very different 
vegetation communities including Harvard Forest (Massachusetts) and San Joaquin 
Experimental Range (California). The data includes spatial and temporal raster 
data ([Link] Module XXX), vector data ([Link] Module XXX), and various 
atmospheric parameters that are collected over several years (this module) that 
she would like to be able explore prior to her initial site visits and writing 
her dissertation proposal. 

 

You are now ready for Lesson 01 in the Tabular Time Series Module

