---
layout: post
title: "Lesson 01: Understanding Time Series Metadata"
date:   2015-10-24
authors: [Megan A. Jones, Marisa Guarinello, Courtney Soderberg]
contributors: [Leah A. Wasser]
dateCreated:  2015-10-22
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
category: 
tags: [module-1]
description: "This lesson teaches basic good practices for writing a script in R
when importing tabular data from a .csv file, how to examine the structure of a 
R object and the importance of collecting information from related metadata 
text files or websites."
code1: 
image:
  feature: NEONCarpentryHeader_2.png
  credit: A collaboration between the National Ecological Observatory Network (NEON) and Data Carpentry
  creditlink: http://www.neoninc.org
permalink: /R/Time-Series-Metadata
comments: false
---

{% include _toc.html %}


##About
This lesson teaches good practices for writing a script in `R` when importing 
tabular data from a .csv file, how to examine the structure of a `R object` and
the importance of collecting information from related metadata text files and
websites.

**R Skill Level:** Intermediate - you've got the basics of `R` down and 
understand the general structure of tabular data.

<div id="objectives" markdown="1">

###Goals / Objectives
After completing this lesson, you will:

* Know how to import a .csv file and examine the structure of the related `R`
object. 
* Use a metadata file to better understand the content of a dataset.
* Understand the importance of including metadata details in your `R` script.
* Know what an EML file is. 

###Things You'll Need To Complete This Lesson
Please be sure you have the most current version of R and, preferably,
RStudio to write your code.

####R Libraries to Install
No additional libraries are needed for this lesson.

####Data to Download
<a href="http://files.figshare.com/2437700/AtmosData.zip" class="btn btn-success">
Download Atmospheric Data</a>

The data used in this lesson were collected at Harvard Forest which is
an National Ecological Observatory Network  
<a href="http://www.neoninc.org/science-design/field-sites/harvard-forest" target="_blank"> field site</a>. These data are proxy data for what will be available for 30 years on the [NEON data portal](http://data.neoninc.org/ "NEON data")
for both Harvard Forest and other field sites located across the United States.

####Setting the Working Directory
The code in this lesson assumes that you have set your working directory to the
location of the unzipped file of data downloaded above.  If you would like a
refresher on setting the working directory, please view the [Setting A Working Directory In R]({{site.baseurl}}/R/Set-Working-Directory "R Working Directory Lesson") lesson prior to beginning
this lesson. 

</div>

#The Data Approach
If our goal were to exploring the relationships between air temperature, 
precipitation,  
photosynthetically active radiation (PAR) and plant phenology we would have to
know how the data were collected and what units the variables were in prior to
any plotting or further analysis. 

##Good practices in R Script Writing
When initiating a new script and workflow it is always a good idea to 

 1) load all necessary packages,
 2) set the working directory,
 3) load the data and
 4) figure out how `R` should interpret that data and how `R` actually interprets
it.  

Let's go through these steps with our data. 

Load any needed packages prior to importing and working with the data.  At this
point we
will only be working with base R and don't need additional packages. If we need 
another package later, we will return to this section to add it so that at a 
glance we can see all the packages in the script.  Including a short note in the 
code allows us to denote each package's purpose in the script. This will help 
later when we write new scripts for other projects.

``` {r load-libraries}
# Load packages required for entire script
# library(PackageName)  #purpose of package
```

Once the libraries are loaded, we need to set the working directory to the
location where our data is stored. 

``` {r set-wd}
#set working directory to ensure R can find the file we wish to import
#setwd("working-dir-path-here")
```

We are now ready to import the data.  This data is a .csv or comma-separated
value file.  This file format is an excellent way to store tabular data in a
plain text format that can be read by many different programs including R. 

We will use the `hf001-10-15min-m.csv` file that contains micro-meteoreology
data including our variables of interest: temperature, precipitation, and PAR 
for Harvard Forest, aggregated at 15 minute intervals.

``` {r load-data}
# Load csv file of 15 min meteorological data from Harvard Forest
#Factors=FALSE so strings, a series of letters/ words/ numerals, remain characters
harMet_15Min <- read.csv(file="AtmosData/HARV/hf001-10-15min-m.csv",
      stringsAsFactors = FALSE)
```

Now it is time to look at our data, metadata and information about how `R` 
interprets the data.

# Data Structure 
We have three basic options for viewing the data structure.  We can view the
first few lines of data using `head()`, we can use `str()` to see how
`R` *interprets* the data or we could view the data in a spreadsheet format by
opening the object using `View()` or clicking on the object in the Environment pane of RStudio.

```{r look-at-data-structure}
#to see first few lines of data file
head(harMet_15Min)

#Check how R is interpreting the data. What is the structure (str) of the data? 
#Is it what we expect to see?  
str(harMet_15Min)

#to see it in spreadsheet form and scroll
View(harMet_15Min)
```

We can see that because we used the `stringsAsFactors=FALSE` command, data types 
that are not numbers or integers are classified as characters (chr) rather than 
as factors.

#Metadata
There is information other than how `R` interprets our data that we need to know.

Can you think of any examples? 

When we are examining data that we've obtained from others, the extra step of
metadata investigation is essential to visualizing, analyzing and interpreting
the data properly for a project.

##Open Associated Metadata Files
To get some of this information, or clues to this information, we will open the
metadata text file associated with this 15-min meteorology data that is in the 
AtmosData folder. Navigate to this file and open it in a text editor.

![Harvard Forest 15 min Atmospheric Data Screenshot]  ({{site.baseurl}}/images/tabTimeSeries/TS01_MetaDataScreenShot.tiff)

Remember, we are currently interested in plant phenology so we'll be looking at
air temperature, precipitation and PAR. Look for these variables and identify
what information is available for each in this metadata file. 

As we find pertinent information about our variables we will add it to the 
script so that when revisiting the script in the future we don't have to look at
the metadata file again. These notes are also helpful if sharing the script with 
others, so we will make notes that are intuitive and intelligible not only to 
ourselves, but also to colleagues and collaborators.

```{r metadata-debrief}
#Metadata Notes from hf001_10-15-m_Metadata.txt
# column names for variables we are going to use: datetime, airt, prec, parr 
# units for quantitative variables: Celsius, millimeters, molePerMeterSquared
# airt and parr are averages of measurements taken every 1 sec; precip is total of each 15 min period 
# for quantitative variables missing values are given as NA
```

## Metadata Sleuthing
Questions: 

* Is there information provided here that doesn't match the data? 
* Is there information we want to know that is not provided in this metadata
file? 
* Do you have ideas how we might find this information?

We notice that no information is given for `datetime` besides that it is a date 
and time. It would be nice to know if it was recorded in UTC/GMT or local time.
We also noticed that there is a link to the source website at the top of the 
text file. Let's check it out to see if we can get the answer to our question.


The website gives more information that we have in our metadata file. And it 
gives us important information about `datetime` -- all values are in Eastern 
Standard Time (see Methods: Observation Periods).

Again, it is a good practice to make relevant notes from the metadata in the 
script so that we can easily refer to it as we actually work with the data.

```{r metadata-website}
# website for more information: http://harvardforest.fas.harvard.edu:8080/exist/apps/datasets/showData.html?id=hf001
# date-times are in Eastern Standard Time
# preview tab give plots of all variables
```

## Other Metadata File Formats
Do you see anything else on the website that might be related to metadata? 

An EML file! This is the official standards-based metadata file using the 
Ecological Metadata Language (EML). Download this to your data folder (EML file:
knb-lter-hfr.1.21). The Harvard Forest is an Long Term Ecological Research
Network (LTER) site and LTER sites use EML extensively in their projects.

```{r metadata-eml-file}
# EML file in the data folder now - all metadata information in there.
```

Now we have important information that will help us as we continue to use these
data for time series analysis.
