---
layout: post
title: "Lesson 05: Plotting with Facets and NDVI Data"
date:   2015-10-20
authors: [Megan A. Jones, Marisa Guarinello, Courtney Soderberg]
contributors: [Leah Wasser]
dateCreated: 2015-10-22
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
category:  
tags: [module-1]
mainTag: 
description: "This lesson will teach individuals how to plot subsetted timeseries data (e.g., plot by season) and to plot time series data with NDVI."
code1:
image:
  feature: NEONCarpentryHeader_2.png
  credit: A collaboration between the National Ecological Observatory Network (NEON) and Data Carpentry
  creditlink: http://www.neoninc.org
permalink: /R/Data-Exploration
comments: false
---

{% include _toc.html %}

##About
This lesson will teach individuals how to plot subsetted timeseries data (e.g., plot by season) and to plot time series data with NDVI.

**R Skill Level:** Intermediate - you've got the basics of `R` down.

<div id="objectives" markdown="1">

###Goals / Objectives
After completing this activity, you will:
 * Know how to use facets in ggplot.
 * Be able to combine different types of data into one plot.

###Things You'll Need To Complete This Lesson
Please be sure you have the most current version of `R` and, preferably,
RStudio to write your code.

####R Libraries to Install
<li><strong>ggplot:</strong> <code> install.packages("ggplot2")</code></li>
<li><strong>scales:</strong> <code> install.packages("scales")</code></li>
<li><strong>gridExtra:</strong> <code> install.packages("gridExtra")</code></li>

####Data to Download
Make sure you have downloaded the Landsat_NDVI folder from
http://figshare.com/articles/NEON_Spatio_Temporal_Teaching_Dataset/1580068

The data used in this tutorial were collected at Harvard Forest which is
a the National Ecological Observatory Network field site <a href="http://www.neoninc.org/science-design/field-sites/harvard-forest" target="_blank">
More about the NEON Harvard Forest field site</a>. These data are proxy data for what will be
available for 30 years from the NEON flux tower [from the NEON data portal](http://data.neoninc.org/ "NEON data").

####Setting the Working Directory
The code in this lesson assumes that you have set your working directory to the
location of the unzipped file of data downloaded above.  If you would like a
refresher on setting the working directory, please view the <a href="XXX" target="_blank">Setting A Working Directory In R</a> lesson prior to beginning the lesson.

####Recommended Pre-Lesson Reading
* Time Series In R: Workshop Introduction Page [LINK]
* Phenology Introduction Page [LINK]
* A brief video explaining <a href="https://www.youtube.com/watch?v=rxOMhQwApMc" target="_blank"> NDVI</a>. 

</div>

In studying phenology it is likely that some combination of precipitation,
temperature, and solar radiation (PAR) plays a role in the annual greening and
browning of plants.  

Let's look at interactions between several of the variables to each other across
the seasons before adding the Normalized Difference Vegetation Index (NDVI) data
into the mix.  Using the NDVI data we can finally directly compare the observed
plant phenology with patterns we've already been exploring.  

##Loading the Data
You will need the daily and micro-meteorology data for 2005-2011 from
the Harvard Forest. If you do not have these data sets, as R objects created in
[{{site.baseurl}} / LINKHERE](Lesson 02 - ), please load them from the .csv files in the downloaded data.  

```{r load-data}
#Remember it is good coding technique to add additional libraries to the top of
  #your script 
library(lubridate) #for working with dates
library(ggplot2)  #for creating graphs
library(scales)   #to access breaks/formatting functions
library(dplyr)    #for data manipulation (split, apply, combine)
library(gridExtra) #for arranging plots

#set working directory to ensure R can find the file we wish to import
#setwd("working-dir-path-here")

#If not already available from Lesson 01-03:
#daily HARV met data
harMet.daily <- read.csv(file="AtmosData/HARV/hf001-06-daily-m.csv",
                     stringsAsFactors = FALSE)

#covert date to POSIXct data type
harMet.daily$date <- as.POSIXct(harMet.daily$date)
```


##Graph precipitation by total PAR across all seasons
PAR, a measure of solar radiation, is less on cloudy days and precipitation is
 more likely when clouds are present, therefore we can expect a relationship
 between them.  We will use `ggplot` to graph PAR
and precipitation from the daily Harvard Meterological data.  We can simply do
this using the code we previously learned and substituting precipitation
(prec) in for time on the x axis.  

```{r PAR-v-precip}
#PAR v precip 
par.precip <- ggplot(harMet.daily,aes(prec, part)) +
           geom_point(na.rm=TRUE) +    #removing the NA values
           ggtitle("Daily Precipitation and PAR at Harvard Forest") +
           theme(plot.title = element_text(lineheight=.8, face="bold",size = 20)) +
           theme(text = element_text(size=20)) +
           xlab("Total Precipitation (mm)") + ylab("Mean Total PAR")
par.precip
```

While there is a lot of noise in the data, there does seem to be a trend of 
lower PAR when precipitation is high. Yet in this data we don't tease apart any 
of annual patterns as date isn't any part of this figure.  

##Use daily data and subset by seasons (retaining PAR, precip, & temp variables)
One way to add a time component to this figure is to subset the data by season 
and then to plot a facetted graph showing the data by season.  

The first step to doing this is to subset by season.  First, we have to switch 
away from coding and data analysis and think about ecology.  What is the best
way to break a year apart into seasons?  

The divisions will change depending on the geographic location and climate where 
the data were collected.  We are using data from Harvard Forest in Massachusetts 
in the northeastern United States.  Given the strong seasonal affects in this 
area dividing the year into 4 seasons is ecologically relevant.  Based on
general knowledge of the area it is also plausible to group the months as 

 * Winter: December - February
 * Spring: March - May 
 * Summer: June - August
 * Fall: September - November 
 
In order to subset the data by season we will use the `dplyr` package.

```{r subsetting-by-season}
#subset by season using month (12-2 is winter, 3-5 spring, 6-8 summer, 9-11 fall)
#add month column
harMet.daily <- harMet.daily  %>% 
  mutate(month=format(date,"%m"))

#check structure of this variable
str(harMet.daily$month)
```

Notice month is a character.   Because month is a character variable we will 
need to put the month number in quotes  when we use mutate to create a new 
variable for season. 

When using `dplyr` %in% is short-hand for -or-, so the 3rd line of code 
essentially says "If the month variable is equal to 12 or to 01 or to 02, set
the season variable to winter".

``` {r subsetting-by-season-2}
harMet.daily <- harMet.daily %>% 
  mutate(season = 
           ifelse(month %in% c("12", "01", "02"), "winter",
           ifelse(month %in% c("03", "04", "05"), "spring",
           ifelse(month %in% c("06", "07", "08"), "summer",
           ifelse(month %in% c("09", "10", "11"), "fall", "Error")))))

#check to see if this worked
head(harMet.daily)
tail(harMet.daily)

```

##Use Facets in ggplot
This allows us to create the same graph for each season and display them in a
grid. Also break up this chunk and move some in code directions to text. 

```{r plot-by-season}

#run this code to plot the same plot as before but with one plot per season
par.precip + facet_grid(. ~ season)
```

Why this didn't work? 

We added the season variable to harMet.daily after we created the original 
`par.precip` plot. Go back up to the code we used to create `par.precip` and run
it again; now it will include the new version of harMet.daily that has the 
season variable.

Let's try the new plot again. 
``` {r plot-by-season2}
par.precip + facet_grid(. ~ season)
# for a landscape orientation of the plots we change the order of arguments in facet_grid():
par.precip + facet_grid(season ~ .)
#and another arrangement of plots:
par.precip + facet_wrap(~season, ncol = 2)

```

#Graph variables and NDVI data together
Add discussion of why you might want to do this. 

```{r NDVI plots}
#first read in the NDVI CSV data
NDVI.2009 <- read.csv(file="Landsat_NDVI/HARV/Harv2009NDVI.csv",
                      stringsAsFactors = FALSE)
#check out the data
str(NDVI.2009)
View(NDVI.2009)

###plot of NDVI vs. PAR using daily data
#first lets get just 2009 from the harmet.Daily data since that is the only year for which we have NDVI data
harMet.daily2009 <- harMet.daily %>% 
  mutate(year = year(date)) %>%   #need to create a year only column first
  filter(year == "2009")

#ggplot does not provide for two y-axes and the scale for these two variables are vastly different.
#So we will create a plot for each variable using the same time variable (julian day) as our x-axis.
#Then we will plot the two plots in the same viewer so we can compare

#create plot of julian day vs. PAR
par.2009 <- ggplot(harMet.daily2009, aes(jd,part))+
  geom_point(na.rm=TRUE)+
  ggtitle("Daily PAR at Harvard Forest, 2009")+
  theme(legend.position = "none", plot.title = element_text(lineheight=.8, face="bold",size = 20)) +
           theme(text = element_text(size=20)) 

#create plot of julian day vs. NDVI
NDVI.2009 <- ggplot(NDVI.2009,aes(julianDays, meanNDVI))+
  geom_point(aes(color = "green", size = 4)) +
  ggtitle("Daily NDVI at Harvard Forest, 2009")+
  theme(legend.position = "none", plot.title = element_text(lineheight=.8, face="bold",size = 20)) +
           theme(text = element_text(size=20)) 

#display the plots together
grid.arrange(par.2009, NDVI.2009) 

#Let's take a look at air temperature too
airt.2009 <- ggplot(harMet.daily2009, aes(jd,airt))+
  geom_point(na.rm=TRUE)+
  ggtitle("Daily Air Temperature at Harvard Forest, 2009")+
  theme(legend.position = "none", plot.title = element_text(lineheight=.8, face="bold",size = 20)) +
           theme(text = element_text(size=20)) 

grid.arrange(airt.2009, NDVI.2009)
#all 3 together
grid.arrange(par.2009, airt.2009, NDVI.2009)

```


#Polishing your ggPlot graphs
Throughout this module we have been creating plots using ggplot.  We've covered
the basics of adding axis labels and titles but using ggplot you can do so much 
more.  If you are interested in making your graphs look better consider
following up with any of these resources:

1) ggplot2 Cheatsheet from Zev Ross: <a href="http://neondataskills.org/cheatsheets/R-GGPLOT2/" target="_blank"> ggplot2 Cheatsheet</a>  
2) ggplot2 documentation index: 
 <a href="http://docs.ggplot2.org/current/index.html#" target="_blank"> ggplot2 Documentation</a>    
3) NEON's Publishable Maps Tutorial ADD LINK TO PRETTY MAPS TUTORIAL WHEN DONE
