---
layout: post
title: "Lesson 05: Plotting Time Series with ggplot in R"
date:   2015-10-20
authors: [Megan A. Jones, Marisa Guarinello, Courtney Soderberg]
contributors: [Leah A. Wasser, Michael Patterson]
dateCreated: 2015-10-22
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
tags: [spatio-temporal, time-series, phenology]
mainTag: time-series
packagesLibraries: [lubridate, ggplot2, dplyr]
category: 
description: "This lesson uses ggplot2 to create professional plots of time 
series data. We will learn how to adjust x and y axis ticks using the scales
package, how to add trend lines to a scatter plot and how to customize plot
labels, colors and overall theme."
code1: TS05-Plotting-Time-Series-ggplot-In-R.R
image:
  feature: NEONCarpentryHeader_2.png
  credit: A collaboration between the National Ecological Observatory Network (NEON) and Data Carpentry
  creditlink: http://www.neoninc.org
permalink: /R/Time-Series-Plot-ggplot
comments: false
---

{% include _toc.html %}


##About
This lesson uses `ggplot2` to create professional plots of time 
series data. We will learn how to adjust x and y axis ticks using the scales
package, how to add trend lines to a scatter plot and how to customize plot
labels, colors and overall theme.

**R Skill Level:** Intermediate - you've got the basics of `R` down.

<div id="objectives" markdown="1">

### Goals / Objectives
After completing this lesson, you will:
 * Be able to create basic time series plots using `ggplot()` in `R`.
 * Understand the synthax of `ggplot()` and know how to find out more about the
 package. 
 * Be able to plot data using scatter and bar plots.
 
**To complete this lesson:** you will need the most current version of R, and 
preferably RStudio, loaded on your computer.

####R Libraries to Install
* **lubridate:** `install.packages("lubridate")`
* **ggplot2:** `install.packages("ggplot2")`
* **scales:** `install.packages("scales")`
* **gridExtra:** `install.packages("gridExtra")`

  <a href="http://neondataskills.org/R/Packages-In-R/" target="_blank">
More on Packages in R - Adapted from Software Carpentry.</a>

####Data to Download
<a href="https://ndownloader.figshare.com/files/3579861" class="btn btn-success"> Download Atmospheric Data</a>

The data used in this lesson were collected at Harvard Forest which is
an National Ecological Observatory Network  
<a href="http://www.neoninc.org/science-design/field-sites/harvard-forest" target="_blank"> field site</a>. 
These data are proxy data for what will be available for 30 years
on the [NEON data portal](http://data.neoninc.org/ "NEON data")
for both Harvard Forest and other field sites located across the United States.

**Set Working Directory:** This lessons assumes that you have set your working 
directory to the location of the downloaded and unzipped data subset. [An overview
of setting the working directory in `R` can be found here.]({{site.baseurl}}/R/Set-Working-Directory "R Working Directory Lesson") 
lesson prior to beginning this lesson.

**Challenge Code:** NEON Data lesson often contain challenges that reinforce 
learned skills. If available, the code for challenge solutions is found in a 
downloadable `R` script available on the footer of each lesson page.


</div>

****

###Additional Resources

 * Winston Chang's <a href="http://www.cookbook-r.com/Graphs/index.html" target="_blank"> Cookbook for R</a> site based on his R Graphics Cookbook text. 
 * The NEON #WorkWithData lesson on <a href="http://neondataskills.org/R/Plotly/" target="_blank"> 
Interactive Data Viz Using R, GGPLOT2 and PLOTLY</a>.
 * Data Carpentry's <a href="http://www.datacarpentry.org/R-ecology/05-visualisation-ggplot2.html" target="_blank"> 
Data Visualization with ggplot2 lesson</a> Data Visualization with ggplot2 lesson.
 * Hadley Wickham's <a href="http://docs.ggplot2.org/" target="_blank"> 
documentation</a> on the `ggplot2` package. 

****

#Plotting Time Series Data
Plotting our data allows us to quickly see general patterns including 
outlier points and trends. Plots are also a useful way to communicate the results
of our research. `GGPLOT2` is a powerful `R` library that we use to create customized,
professional plots.

###Load the Data
We will use the `lubridate`, `ggplot2`, `scales` and `gridExtra` packages in this 
lesson.

Our data sample will be the daily meteorology data for 2009-2011 for the NEON
Harvard Forest field site (`AtmosData/HARV/Met_HARV_Daily_2009_2011.csv`). If this
subset is not already loaded in RStudio, please load it now.  

```{r load-data}
#Remember it is good coding technique to add additional libraries to the top of
  #your script 
library(lubridate) #for working with dates
library(ggplot2)  #for creating graphs
library(scales)   #to access breaks/formatting functions
library(gridExtra) #for arranging plots

#set working directory to ensure R can find the file we wish to import
#setwd("working-dir-path-here")

#daily HARV met data, 2009-2011
harMetDaily.09.11 <- read.csv(file="NEON-DS-Met-Time-Series/HARV/Met_HARV_Daily_2009_2011.csv",
                     stringsAsFactors = FALSE)

#covert date to POSIXct date-time class
harMetDaily.09.11$date <- as.POSIXct(harMetDaily.09.11$date)

#monthly HARV temperature data, 2009-2011
harTemp.monthly.09.11<-read.csv(file=
                        "NEON-DS-Met-Time-Series/HARV/Temp_HARV_Monthly_09_11.csv",
                        stringsAsFactors=FALSE)
#convert datetime from chr to datetime class & rename date for clarification
harTemp.monthly.09.11$date <- as.POSIXct(harTemp.monthly.09.11$datetime)
```


##Plot with qplot
We can use the `qplot()` function in the `ggplot2` package to quickly plot a variable
such as air temperature (`airt`) across all three years of our daily averaged 
time series data. 

``` {r qplot}
qplot(x=date, y=airt,
      data=harMetDaily.09.11, na.rm=TRUE,
      main="Air temperature Harvard Forest\n 2009-2011",
      xlab="Date", ylab="Temperature (°C)")
```

The resulting plot displays the pattern of air temperature increasing and 
decreasing over three years. While `qplot()` is a quick way to plot data, our 
ability to customize the output plot is limited.

##Plot with ggplot
The `ggplot()` function within `ggplot2` package gives us more control
over plot appearance. However, to use `ggplot` we need to learn a slightly 
different syntax. Three basic elements are needed:

 1. The **data_frame:** containing the variables that we wish to plot,
 2. **aes (aesthetics):** which denotes which variables will map to the x-, y- (and 
other) axes,  
 3. `geom_XXX` (geometry) which defines the data's graphical representation (e.g.
 points (`geom_point`), bars (`geom_bar`), lines (`geom_line`), etc).
 
The syntax begins with the base statement that includes the data.frame (`harMetDaily.09.11`) 
and associated x (`date`) and y (`airt`) variables to be plotted:

`ggplot(harMetDaily.09.11, aes(date, airt))`

To successfully plot, the last piece that is needed is the `geom`etry type. In 
this case, we want to create a scatterplot so we can add ` + geom_point()`.

Let's create an air temperature scatterplot. 

``` {r basic-ggplot2}
#plot Air Temperature Data across 2009-2011 using 15-minute data
ggplot(harMetDaily.09.11, aes(date, airt)) +
           geom_point(na.rm=TRUE)

```

##Customize A Scatterplot
We can customize our plot in many ways. For instance, we can change the size and 
color of the points using `size=`, shape `pch=`, and `color=` in the geom_point 
element. 

`geom_point(na.rm=TRUE, color="blue", size=1)`

``` {r basic-ggplot2-colors}
#plot Air Temperature Data across 2009-2011 using 15-minute data
ggplot(harMetDaily.09.11, aes(date, airt)) +
           geom_point(na.rm=TRUE, color="blue", size=3, pch=18)

```

#Modify Title & Axis Labels

We can modify plot attributes by adding elements using the `+` symbol.
For example, we can add a title by using `+ ggtitle="XXX",` and axis
labels using `+ xlab("XXX") + ylab("XXX")`.  

``` {r basic-ggplot2-labels }
#plot Air Temperature Data across 2009-2011 using 15-minute data
ggplot(harMetDaily.09.11, aes(date, airt)) +
           geom_point(na.rm=TRUE, color="blue", size=1) + 
           ggtitle("Air Temperature 2009-2011\n NEON Harvard Forest Field Site") +
           xlab("Date") + ylab("Air Temperature (C)")

```

<i class="fa fa-star"></i> **Data Tip:**  Use `help(ggplot2)` to review the many
elements that can be defined and added to a GGPLOT2 plot.
{: .notice }

##Naming plot objects
We can create a `ggplot` object by assigning our plot to an object name.
When we do this, the plot will not render automatically. To render the plot, we
need to call it in the code.

Assigning plots to an `R` object allows us to effectively add on to, 
and modify the plot later. Let's create a new plot and call it `AirTempDaily`.

``` {r basic-ggplot2-labels-named }
#plot Air Temperature Data across 2009-2011 using 15-minute data
AirTempDaily <- ggplot(harMetDaily.09.11, aes(date, airt)) +
           geom_point(na.rm=TRUE, color="purple", size=1) + 
           ggtitle("Air Temperature 2009-2011\n NEON Harvard Forest Field Site") +
           xlab("Date") + ylab("Air Temperature (C)")

#render the plot
AirTempDaily

```

##Format Dates in Axis Labels

We can adjust the display format (e.g. 2009-07 vs. Jul 09) and the number of 
major and minor ticks for axis date values using `scale_x_datetime`. Let's format 
the axis ticks so they read "month year" (`%b %y`). To do this, we will use the syntax:

`scale_x_datetime(labels=date_format("%b %y")`

Rather than re-coding the entire plot, we can add the `scale_x_datetime` element
to the plot object `AirTempDaily` that we just created. 

```{r format-x-axis-labels }
#format x-axis: dates
AirTempDailyb <- AirTempDaily + 
  (scale_x_datetime(labels=date_format("%b %y")))

AirTempDailyb
```

###Adjust Date Ticks

We can adjust the date ticks too. In this instance, having 1 tick per year may be
enough. If we have the `scales` package loaded, we can use 
`breaks=date_breaks("1 year")` to create a tick for every year.
We can adjust this as needed (e.g. 10 days, 30 days, 1 month, etc).

```{r format-x-axis-label-ticks }
#format x-axis: dates
AirTempDailyb <- AirTempDaily + 
    (scale_x_datetime(breaks=date_breaks("6 months"),
      labels=date_format("%b %y")))

AirTempDailyb

#format x-axis: dates
AirTempDailyb <- AirTempDaily + 
    (scale_x_datetime(breaks=date_breaks("1 year"),
      labels=date_format("%b %y")))

AirTempDailyb


```


<i class="fa fa-star"></i> **Data Tip:**  We can adjust the tick spacing and format
for x and y axes using `scale_x_continuous` or `scale_y_continuous` to format 
a continue variable. Check out `??scale_x_` (tab complete to view the various x 
and y scale options)
{: .notice }

##GGPLOT - Subset by Time

Sometimes we want to zoom into a region of our plot without subsetting the 
entire `data_frame`. To do this, we can define start and end times. We can then
define the `limits` in the `scale_x_datetime` object as follows:

`scale_x_datetime(limits=start.end) +`


```{r subset-ggplot-time }

#Define Start and end times for the subset as R objects that are the time class
startTime <- as.POSIXct("2011-01-01 00:00:00")
endTime <- as.POSIXct("2012-01-01 00:00:00")

#create a start and end time R object
start.end <- c(startTime,endTime)
start.end

#View data for 2011 only
AirTempDailyb <- AirTempDaily + 
    (scale_x_datetime(limits=start.end,
                      breaks=date_breaks("1 year"),
                      labels=date_format("%b %y")))

AirTempDailyb

```

##GGPLOT Themes

We can use the `theme()` element to adjust figure elements.
There are some nice pre-defined themes that we can use as a starting place.

```{r nice-font}
#Apply a black and white stock GGPLOT theme
#note the background is white rather than grey.
AirTempDaily<-AirTempDailyb +
  theme_bw()

AirTempDaily
```

###OPTIONAL BONUS: Import New Themes
There are externally developed themes built by the `R` community
that are worth mentioning. Feel free to experiment with the code below to
install `ggthemes`.

```{r install-new-themes }
#install additional themes
#install.packages('ggthemes', dependencies = TRUE)
library(ggthemes)
AirTempDaily<-AirTempDailyb +
  theme_economist()

AirTempDaily


AirTempDaily<-AirTempDailyb +
  theme_stata()

AirTempDaily

```


###More on Themes:

* <a href="http://docs.ggplot2.org/0.9.2.1/theme.html" target="_blank"> 
Hadley Wickham's documentation.</a>
* <a href="http://zevross.com/blog/2014/08/04/beautiful-plotting-in-r-a-ggplot2-cheatsheet-3/#use-a-new-theme-theme_xx" target="_blank">Zev Ross on themes.</a>
*  <a href="https://github.com/jrnold/ggthemes" target="_blank">A list of themes loaded
in the ggthemes library is found here</a>

##Creating Custom Themes
We can customize theme elements manually too. Let's customize the font size and 
style. 



```{r increase-font-size }
#format x axis with dates
AirTempDaily<-AirTempDailyb +
  #theme(plot.title) allows to format the Title seperately from other text
  theme(plot.title = element_text(lineheight=.8, face="bold",size = 20)) +
  #theme(text) will format all text that isn't specifically formatted elsewhere
  theme(text = element_text(size=18)) 

AirTempDaily

```

<div id="challenge" markdown="1">

##Challenge: Plot Total Daily Precipitation

Create a plot of total daily precipitation using data in the `harMetDaily.09.11` 
`data_frame`. 

* Format the dates on the x-axis: "Month-Year".  
* Create a plot object called 'PrecipDaily".
* Be sure to add an appropriate title in addition to x and y axis labels.
* Increase the font size of the plot text and adjust the number of ticks on the x-axis.

</div>

``` {r challenge-code-ggplot-precip, echo=FALSE}
PrecipDaily <- ggplot(harMetDaily.09.11, aes(date, prec)) +
           geom_point() +
           ggtitle("Daily Precipitation Harvard Forest\n 2009-2011") +
            xlab("Date") + ylab("Precipitation (mm)") +
            scale_x_datetime(labels=date_format ("%m-%y"))+
           theme(plot.title = element_text(lineheight=.8, face="bold",
                 size = 20)) +
           theme(text = element_text(size=18))

PrecipDaily
```

##Figures with Bars

We can use ggplot to create bar plots too. Let's create a bar plot of total daily
precipitation next. A bar plot might be a better way to represent a total daily 
value. To create a bar plot, we change the `geom` element from `geom_point()` to 
`geom_bar()`.  

The default setting for a ggplot bar plot -  `geom_bar()` - is a histogram designated
by `stat="bin"`. However, in this case, we want to plot actual precipitation values.
We can use `geom_bar(stat="identity")` to force ggplot to plot actual values:


``` {r ggplot-geom_bar }
PrecipDailyBarA <- ggplot(harMetDaily.09.11, aes(date, prec)) +
           geom_bar(stat="identity", na.rm = TRUE) +
           ggtitle("Daily Precipitation\n Harvard Forest") +
            xlab("Date") + ylab("Precipitation (mm)") +
            scale_x_datetime(labels=date_format ("%b-%y"),breaks=date_breaks("1 year")) +
            theme(plot.title = element_text(lineheight=.8, face="bold",
                 size = 20)) +
           theme(text = element_text(size=18))

PrecipDailyBarA
```

Note that some of the bars in the resulting plot appear grey rather than black.
This is because R will do it's best to adjust colors of bars that are closely spaced
to improve readibility. If you zoom into the plot, all of the bars are black.

<div id="challenge" markdown="1">
##Challenge

Plot `PrecipDailyBarA` - however just plot the data from 2011 using the `start.end`
object. HINT: you do not need to rebuild the plot. Instead, you can call the 
`PrecipDailyBarA` object and add the scale element (`scale_x_datetime(limits=start.end)`).

</div>

``` {r ggplot-geom_bar-subset, echo=FALSE, warning=FALSE }

#plot year subset
PrecipDailyBarA + 
            scale_x_datetime(limits=start.end)

```

##Color
We can change the bar fill color by specifying color within the
`geom_bar(colour="blue")` element. We can also specify a separate fill and line
color using `fill=` and `line=`.  Colors can be specified by name or hexidecimal 
color codes (e.g, #FF9999). There are many color cheatsheets out there to help 
with color selection!

<a href="http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf" target="_blank">
An R color cheatsheet</a>

``` {r ggplot-color }
#specifying color by name
PrecipDailyBarB <- PrecipDailyBarA+
  geom_bar(stat="identity", colour="darkblue")

PrecipDailyBarB


```


<i class="fa fa-star"></i> **Data Tip:**  For more information on color, including 
color blind friendly color palletes, checkout the 
<a href="http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/" target="_blank">
ggplot2 color information</a> from Winston Chang's *Cookbook* *for* *R* site 
based on the _R_ _Graphics_ _Cookbook_ text. 
{: .notice }
 

##Figures with Lines

We can create line plots too using GGPLOT. To do this, we use `geom_line()` instead
of `bar` or `point`.

```{r ggplot-geom_lines}
AirTempDailyline <- ggplot(harMetDaily.09.11, aes(date, airt)) +
           geom_line(na.rm=TRUE) +  
           ggtitle("Air Temperature Harvard Forest\n 2009-2011") +
           xlab("Date") + ylab("Air Temperature (C)") +
          theme(plot.title = element_text(lineheight=.8, face="bold", 
                                          size = 20)) +
           theme(text = element_text(size=18))
AirTempDailyline
```

Note that lines may not be the best way to represent air temperature data given
lines suggest that the connecting points are directly related. It is important
to consider what type of plot best represents the type of data that you are
presenting.

<div id="challenge" markdown="1">
###Challenge: Combine Points & Lines

You can combine geometries within one plot. For example, you can have a `geom_line()` 
and `geom_point` element in a plot. Add `geom_line(na.rm=TRUE)` to the `AirTempDaily` 
plot. What happens?

</div>

```{r challenge-code-geom_lines&points, echo=FALSE }
AirTempDaily + geom_line(na.rm=TRUE) 
```


##Trend Lines

We can add a trend line, which is a statistical transormation of our data to represent
general patterns, using `stat_smooth()`. `stat_smooth()` 
requires a statistical method as follows:


* For data with < 1000 observations: the default model is loess (a non-parametric 
regression model`
* for data with > 1,000 observations: the default model is or gam (a general additive 
model. 
* We can specify a method such as linear regression (`method="lm"`). 

For this lesson, we will use the default trend line model. The `gam` method will 
be used withgiven we have 1,095 measurements.

Remember a trend line is a statistical transformation of the data, so prior to
adding the line one must understand if a particular statistical transformation
is appropriate for the data.  
{: .notice } 


``` {r ggplot-trend-line}
#adding on a trend lin using loess
AirTempDailyTrend <- AirTempDaily + stat_smooth(colour="green")

AirTempDailyTrend
```


<div id="challenge" markdown="1">
#Challenge: A Trend in Percipitation? 

Create a bar plot of total daily precipitation. Add a:

* Trend line for total daily precipitation. 
* Make the bars purple (or your favorite color!) 
* Make the trend line grey (or your other favorite color). 
* Adjust the tick spacing and format the dates to
appear as Jan 2009
* Render the title in *italics*.

</div>

``` {r challenge-code-linear-trend, echo=FALSE}
ggplot(harMetDaily.09.11, aes(date, prec)) +
      geom_bar(stat="identity", colour="darkorchid4") + #dark orchid 4 = #68228B
      ggtitle("Daily Precipitation with Linear Trend\n Harvard Forest") +
      xlab("Date") + ylab("Precipitation (mm)") +
      scale_x_datetime(labels=date_format ("%b %y"))+
      theme(plot.title = element_text(lineheight=.8, face="italic",
                 size = 20)) +
      theme(text = element_text(size=18))+
      stat_smooth(method="lm", colour="grey")
```

<div id="challenge" markdown="1">
##Challenge: Plot Monthly Air Temperature

Use the `harTemp.monthly.09.11` data_frame to plot the monthly air temperature
across 2009-2011.Name your plot "AirTempMonthly". Be sure to label axes and adjust
the plot ticks as you see fit.

</div>

``` {r plot-airtemp-Monthly, echo=FALSE}
AirTempMonthly <- ggplot(harTemp.monthly.09.11, aes(date, mean_airt)) +
           geom_point() +
           ggtitle("Average Monthly Air Temperature\n Harvard Forest") +
           theme(plot.title = element_text(lineheight=.8, face="bold",
                size = 20)) +
           theme(text = element_text(size=18)) +
           xlab("Date") + ylab("Air Temperature (C)") +
          scale_x_datetime(labels=date_format ("%b%y"))
AirTempMonthly

```

##Displaying Multiple Figures in Same Panel

It is often useful to arrange plots in a panel rather than displaying them 
individually. In base `R`, we use `par(mfrow=())` to accomplish this. However the 
`grid.arrange()` function from the `gridExtra` package provides a more efficient 
approach! 

`grid.arrange` requires 2 things:
1. the names of the plots that you wish to render in the panel.
2. the number of columns  (`ncol`).

`grid.arrange(plotOne, plotTwo, ncol=1)`

Let's plot `AirTempMonthly` and `AirTempDaily` on top of each other. To do this,
we'll specify 1 column..


```{r compare-precip }
#note - be sure library(gridExtra) is loaded!
#stack plots in one column 
grid.arrange(AirTempDaily, AirTempMonthly, ncol=1)
```

<div id="challenge" markdown="1">
#Challenge: Create Panel of Plots

Plot `AirTempMonthly` and `AirTempDaily` next to each other rather than
stacked on top of each other.  

</div>
```{r challenge-code-grid-arrange, echo=FALSE}
grid.arrange(AirTempDaily, AirTempMonthly, ncol=2)
```

#Refine ggplot Figures

In this lesson, we've covered the basics of `ggplot`. There are many great 
resources the cover refining `ggplot` figures. A few are below:

1. ggplot2 Cheatsheet from Zev Ross: <a href="http://neondataskills.org/cheatsheets/R-GGPLOT2/" target="_blank"> ggplot2 Cheatsheet</a>  
2. ggplot2 documentation index: 
 <a href="http://docs.ggplot2.org/current/index.html#" target="_blank"> ggplot2 Documentation</a>    
3. NEON's Publishable Maps Tutorial (Currently in draft, link will be updated
when published)
