---
layout: post
title: "Time Series Culmination Activity: Plot using Facets & Plot NDVI with Time Series Data"
date:   2015-10-16
authors: [Megan A. Jones, Leah A. Wasser]
contributors: []
dateCreated: 2015-10-22
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
packagesLibraries: [ggplot2, scales, gridExtra, grid, dplyr, reshape2]
categories: [self-paced-tutorial]
mainTag: tabular-time-series
tags: [time-series, phenology, R]
tutorialSeries: [tabular-time-series]
description: "This tutorial is a data integration wrap-up culmination activity 
for the spatio-temporal time series tutorials."
code1: TSCulmination-Work-With-NDVI-and-Met-Data-In-R.R
image:
  feature: NEONCarpentryHeader_2.png
  credit: A collaboration between the National Ecological Observatory Network (NEON) and Data Carpentry
  creditlink: http://www.neoninc.org
permalink: R/Work-With-NDVI-and-Met-Data-In-R
comments: false
---

{% include _toc.html %}

## About
This tutorial is a culmination activity for the end of the lesson series on 
[tabular time series data in R ]({{ site.baseurl }}self-paced-tutorials/tabular-time-series) 
and as part of a 
[spatio-temporal Data Carpentry workshop ]({{ site.baseurl }}self-paced-tutorials/spatio-temporal-workshop). 
The data used in this culmination activity has been extracted or previously
worked with in the 
[tabular time series data in R ]({{ site.baseurl }}self-paced-tutorials/tabular-time-series)
and  
[raster data in R ]({{ site.baseurl }}self-paced-tutorials/spatial-raster-series).

**R Skill Level:** Advanced/ Intermediate - you've got the basics of `R` down
 and are comfortable using `ggplot2` and `dplyr`. 

<div id="objectives" markdown="1">

# Goals / Objectives
After completing this activity, you will:

 * 

## Things You’ll Need To Complete This Lesson
To complete this lesson: you will need the most current version of R, and 
preferably RStudio, loaded on your computer.

###Install R Packages
* **ggplot2:** `install.packages("ggplot2")`
* **scales:** `install.packages("scales")`
* **gridExtra:** `install.packages("gridExtra")`
* **grid:** `install.packages("grid")`
* **dplyr:** `install.packages("dplyr")`
* **reshape2:** `install.packages("reshape2")`

[More on Packages in R - Adapted from Software Carpentry.]({{site.baseurl}}R/Packages-In-R/)

### Download Data 
{% include/dataSubsets/_data_Met-Time-Series.html %}

****

{% include/_greyBox-wd-rscript.html %}
{% include/tutorialSeries/_series_dc-tabular-time-series.html %}

****

### Recommended Tutorials

This lessons uses both `dplyr` and `ggplot2`. We recommend the following tutorials
before working through this one, if you are new to either of the above `R` packages.

* [Subset & Manipulate Time Series Data with dplyr lesson]({{site.baseurl}}/R/Time-Series-Subset-dplyr/ "Learn dplyr") 

* [Plotting Time Series with ggplot in R lesson]({{site.baseurl}}/R/Time-Series-Plot-ggplot/ "Learn ggplot")  

</div>

## Plot NDVI & PAR using Daily Data

### NDVI Data
Normalized Difference Vegetation Index (NDVI) is an indicator of how green
vegetation is.  NDVI is derived from remote sensing data based on a ratio the
reluctance of visible red spectra and near-infrared spectra.  The NDVI values
vary from -1.0 to 1.0.  

The imagery data used to create this NDVI data were collected over the National
Ecological Observatory Network's
<a href="http://www.neoninc.org/science-design/field-sites/harvard-forest" target="_blank" >Harvard Forest</a>
field site.  
The imagery was created by the U.S. Geological Survey (USGS) using a 
<a href="http://eros.usgs.gov/#/Find_Data/Products_and_Data_Available/MSS" target="_blank" >  multispectral scanner</a>
on a
<a href="http://landsat.usgs.gov" target="_blank" > Landsat Satellite </a>.
The data files are Geographic Tagged Image-File Format (GeoTIFF).  
A tutorial, [Extract NDVI Summary Values from a Raster Time Series]({{ site.baseurl}}/R/Extract-NDVI-From-Rasters-In-R/), 
explains how to create this NDVI file from on Raster data. 

### Read In NDVI Data
We need to read in the 2011 NDVI data for the Harvard Forest. A .csv file is
within `HARV/NDVI` subdirectories in the `NEON-DS-Met-Time-Series` directory.

```{r read-NDVI}
#Remember it is good coding technique to add additional libraries to the top of
  #your script 
library(lubridate) #for working with dates
library(ggplot2)  #for creating graphs
library(scales)   #to access breaks/formatting functions
library(gridExtra) #for arranging plots
library(grid)   #for arrangeing plots
library(dplyr)  #for subsetting by season

#set working directory to ensure R can find the file we wish to import
#setwd("working-dir-path-here")

#first read in the NDVI CSV data
NDVI.2011 <- read.csv(
  file="NEON-DS-Met-Time-Series/HARV/NDVI/meanNDVI_HARV_2011.csv", 
  stringsAsFactors = FALSE
  )

#check out the data
str(NDVI.2011)
head(NDVI.2011)

```

<div id="challenge" markdown="1">
## Challenge:  Class Conversion & Subset by Time

1. `NDVI.2001$Date` imported into `R` as a character class. Convert the data to 
a date class. 

2. Subset just the data from 2011 from the `harmet.Daily` data. 
</div>

``` {r challenge-code-convert-date, include=TRUE, results="hide", echo=FALSE}
#convert chr class Date to date class Date
NDVI.2011$Date<- as.Date(NDVI.2011$Date)
#double check
str(NDVI.2011)

##2
#Use dplyr to subset only 2011 data
harMet.daily2011 <- harMetDaily.09.11 %>% 
  mutate(year = year(date)) %>%   #need to create a year only column first
  filter(year == "2011")

#convert data from POSIX class to Date class; both "date" vars. now Date class
harMet.daily2011$date<-as.Date(harMet.daily2011$date)
```

In this dataset, we have the following variables:  

* 'X': an integer representing each row
* meanNDVI: the daily total NDVI for that area.  ("Mean" comes from the fact
   that the value is a mean of all pixels in the original raster).
* site: all NDVI values are from the Harvard Forest
* year: all values are from 2011
* julianDay: the numeric day of the year
* Date: a date in format "YYYY-MM-DD".

## Two y-axes or Side-by-Side Plots?
When we have different types of data like NDVI (scale: 0-1 index units) and PAR
(scale: 0-65.8 mole per meter squared) that we want to plot over time, we cannot
simply plot them on the same plot as they have different y-axes.  

One option, would be to plot both data types in the same plot space but each
having it's own axis.  However, there is a line of graphical representation 
thought that this is not a good practice.  The creator of `ggplot2` ascribes to
this dislike of different y-axes and so neither `qplot` nor `ggplot` have this
functionality.  

Instead, plots of different types of data can be plotted next to each other to 
allow for comparison.  Depending on how the plots are being viewed they can have
a vertical or horizontal arrangement.  

Following this second option, we will create a plot for each variable using the
same time variable (Julian day) as our x-axis.

Then we will plot the two plots in the same viewer so we can more easily compare
them.  

``` {r plot-PAR-NDVI}
#create plot of julian day vs. PAR
plot.par.2011 <- ggplot(harMet.daily2011, aes(date, part))+
  geom_point(na.rm=TRUE) +
  ggtitle("Daily PAR at Harvard Forest, 2011")+
  theme(legend.position = "none",
        plot.title = element_text(lineheight=.8, face="bold",size = 20),
        text = element_text(size=20))

plot.NDVI.2011 <- ggplot(NDVI.2011, aes(Date, meanNDVI))+
  geom_point(colour = "forestgreen", size = 4) +
  ggtitle("Daily NDVI at Harvard Forest, 2011")+
  theme(legend.position = "none",
        plot.title = element_text(lineheight=.8, face="bold",size = 20),
        text = element_text(size=20))
 
#display the plots together
grid.arrange(plot.par.2011, plot.NDVI.2011) 
```

This is nice but a bit confusing as the time on our x-axis doesn't fully line
up.  To fix this we can assign the same min and max to both x-axes so that they
align.  We can also label the axes. 

``` {r plot-same-xaxis}
plot2.par.2011 <- plot.par.2011 +
               scale_x_date(labels = date_format("%b %d"),
               date_breaks = "3 months",
               date_minor_breaks= "1 week",
               limits=c(min=min(NDVI.2011$Date),max=max(NDVI.2011$Date))) +
                ylab("Total PAR") + xlab ("")

plot2.NDVI.2011 <- plot.NDVI.2011 +
          scale_x_date(labels = date_format("%b %d"),
               date_breaks = "3 months", 
               date_minor_breaks= "1 week",
               limits=c(min=min(NDVI.2011$Date),max=max(NDVI.2011$Date)))+
  ylab("Total NDVI") + xlab ("Date")

grid.arrange(plot2.par.2011, plot2.NDVI.2011) 

```

<div id="challenge" markdown="1">
## Challenge: Plot Air Temperature and NDVI
Create a complementary plot pairing with Air Temperature and NDVI.  Choose 
colors and symbols that show the data well.  Finally, plot air temperature, PAR,
and NDVI in a single window to ease comparisons.  
</div>

``` {r challengeplot-same-xaxis, echo=FALSE}
plot.airt.2011 <- ggplot(harMet.daily2011, aes(date, airt))+
  geom_point(colour="darkblue", na.rm=TRUE) +
  ggtitle("Average Air Temperature\n Harvard Forest 2011")+
  scale_x_date(labels = date_format("%b %d"),
               date_breaks = "3 months", date_minor_breaks= "1 week",
               limits=c(min(NDVI.2011$Date),max=max(NDVI.2011$Date)))+
  ylab("Celcius") + xlab ("")+
  theme(legend.position = "none",
        plot.title = element_text(lineheight=.8, face="bold",size = 20),
        text = element_text(size=20))

grid.arrange(plot.airt.2011, plot2.NDVI.2011) 

grid.arrange(plot2.par.2011, plot.airt.2011, plot2.NDVI.2011) 
```

## Two Plots with One x-axis.  
We are able to nicely see the three different variables.  However, we waste a
lot of space to repeating the x-axes and the titles.  Instead of three separate
plots we could alternatively, use facets and plot all the variables of interests
(NDVI, air temperature, precipitation, and PAR) on a single x-axis.  To do this
we'll use `melt()` from the `reshape2` package. `melt` transforms the data from
a wide format (columns of different variables with values in them), to a long
format (column of variables with a column of values) organized according to a
specified variable.  

```{r plot-same-x-axis-1}
library(reshape2)  #allows us to "melt" dataframes from "wide" to "long"

#merge the two data frames by date and retain all 'harMet.daily
harMetNDVIall.daily.2011<- merge(harMet.daily2011, 
                                 NDVI.2011, by.x = "date", 
                                 by.y = "Date", all.x=TRUE)

#convert from "wide" form to "long" form
harMetNDVI.daily.2011.long<-melt(harMetNDVIall.daily.2011, id ="date")
```

Once we have the data in the long form we can subset and then plot the data. 

``` {r plot-same-x-axis-2}
#subset to retain just the variables of interest.  The vertical bar character
# means "OR".
harMetNDVI.daily.2011.select<-subset(harMetNDVI.daily.2011.long,
                                variable=="meanNDVI"| variable== "prec"|
                                variable == "airt" | variable == "part")

NDVI.harMet.facet.plot<-ggplot(
  harMetNDVI.daily.2011.select,
  aes(date, value), group=variable) +
  geom_point() +
  facet_grid(variable~., scales="free") +   #specify facets & y-axis can vary
  ggtitle("Harvard Forest 2011") +
  scale_x_date(labels = date_format("%b %d"),  #abbreviated month & day
               date_breaks = "3 months", date_minor_breaks= "1 month") +  #where grid is
  xlab ("Date") + ylab ("Value") +
  theme(legend.position = "none",
        plot.title = element_text(lineheight=.8, face="bold",size = 20),
        text = element_text(size=10)
  ) 

NDVI.harMet.facet.plot

```
